Parallelizing Linear Recurrent Neural Nets Over Sequence Length

### 2017-10-27 paper/main.tex
Final submission material

We investigate this idea further and deliver the following contributions:
\begin{itemize}
\item{We classify RNNs which satisfy the conditions above, and show that many
    RNNs used in practice such as the Quasi-RNNs (QRNNs) introduced by
    \citet{bradbury2017quasi} are contained in this class.}
\item{We provide an implementation of the parallel linear recurrence algorithm
    as a CUDA kernel, and show that it speeds up training of QRNN and
    \citet{lei2017}'s Simple Recurrent Unit (SRU) architectures by factors of up
    to 9x.}
\item{We describe how several recent linear RNNs can be described as
    linear surrogates for non-linear architectures. We introduce a linear
    surrogate for the LSTM and show that we are able to train it
    with a speedup of 5-10x compared to the CuDNN-LSTM when we use the
    parallel linear recurrence algorithm.}
\end{itemize}

@@

\subsection{Synthetic Experiment}
One of the key strengths of the LSTM is that it is capable of dealing with
long-term dependencies. In order to demonstrate that the GILR-LSTM is also able
to handle long-term dependencies we tackle a canonical example of inference over
many time steps from \citet{hochreiter1997long}.  We show that in fact the
GILR-LSTM is able to outperform the CuDNN-LSTM and extend to sequence lengths
orders of magnitude longer than dealt with previously. The input consists of
sequences of length $n$ where for $n > 0$ each element is a randomly chosen
one-hot vector \(x\) in $p$-dimensional space.  The first vector in each
sequence, \(x_0\), is always either \((1, 0, \ldots, 0)\) or \((-1, 0, \ldots,
0)\). The sequential model must read in an entire sequence and then output the
sign of the first sequence element.  This sequence classification problem
requires remembering the first element over the length of the sequence, and
early RNNs struggled with this for \(p\) as small as a few dozen. In the
original formulation of the problem (dealing in the regime with around one
hundred timesteps), the dimensionality of the input \(p\) is set equal to
\(n\). Since this would make the size of the input data grow impractically large
as \(\mathcal{O}(n^2)\) for long sequences, we fix \(p = 128\) as we vary \(n\).
\begin{figure}[] \centering
\includegraphics[width=12cm]{synthetic_diagram.pdf}
\caption{The structure of the synthetic example and the GILR-LSTM architecture
we used to tackle it. We feed in one-hot unit vectors \(x\) which are chosen
uniformly at random (with replacement). The class is determined by the very
first vector \(x_0\), which has a fixed direction. The sign of \(x_0\)
determines the class. In the diagram, each rounded block indicates a cell of the
RNN, whilst the square indicates a linear unit.}
\label{fig:synthetic_diagram}
\end{figure} We generated sequences for $n$ equal to 1,024, 8,192, and
1,048,576. For each of these we compared a two layer GILR-LSTM with 512 hidden
units to a two layer LSTM with 512 hidden units\footnote{For the longest
sequence length, the number of hidden units was decreased to 64 for both
architectures so that the net could fit in memory.} per layer implemented by
CuDNN.

@@

\begin{table}[]
\centering
\caption{Performance of the GILR-LSTM compared to the CuDNN LSTM
  on problem 2b from \citet{hochreiter1997long}. }
\label{table:synth-table}
\begin{tabular}{@{}lllllll@{}} \toprule Sequence Length &
\multicolumn{2}{c}{\textbf{1,024}} & \multicolumn{2}{c}{\textbf{8,192}} &
\multicolumn{2}{c}{\textbf{1,048,576}} \\ \midrule &CuDNN& GILR &
CuDNN &GILR&CuDNN&GILR\\ \cmidrule(l){2-7} Iterations (1000s) &
                     1.0 %bs=8
                     \(\pm\) 0.4 & 0.55 %bs=8
                                    \(\pm\) 0.04 & 0.44 %bs=8
                                                   \(\pm\) 0.05 & 0.56 %bs=8
                                                                 \(\pm\) 0.16
                                                               &-& 14 \(\pm\) 3 \\
\begin{tabular}[c]{@{}l@{}}Wall time (hours)\end{tabular} &
0.28 \(\pm\) 0.08 & 0.031 \(\pm\) 0.002 & 0.58 \(\pm\) 0.06 & 0.10 \(\pm\) 0.03
&-& 9.7 \(\pm\) 1.7 \\ \bottomrule
\end{tabular}
\end{table}

The results as show in table \ref{synth-table} illustrate that the GILR-LSTM is able to converge between 6 and 10
times faster than the CuDNN-LSTM. This is somewhat surprising
given the LSTM was specifically constructed for problems of this sort, and the
CuDNN LSTM implementation is highly optimized (to the extent that
the monolithic interface it exposes is difficult to modify or extend). The
GILR-LSTM is implemented entirely in standard TensorFlow with the exception of
the call to new linear recurrence kernel in place of a TensorFlow symbolic
loop.
Convergence of the GILR-LSTM models
leads to the conclusion that the non-linearities present in LSTM are
not necessary for solving this instance of the long-term dependency problem.
The time to convergence further leads to the conclusion that inclusion of a
non-linearity at every step incurs a significant training time slowdown.
Furthermore, the GILR-LSTM is able to learn to carry dependencies over a one
million element sequence. As far as we know, this one million step sequence
experiment is the longest sequential learning problem to be handled by neural
networks to date.

\begin{figure}
  \centering
  \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./1k_synthetic_new.png}
  \end{minipage}%
    \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./8k_synthetic_new.png}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./1m_synthetic_new.png}
  \end{minipage}
  \caption{Training curves for GILR-LSTM and CuDNN-LSTM architectures for various
    sequence lengths. Since we used
    small batch sizes to train and evaluate accuracy, the accuracy fluctuated
    (between 0 and 1 for the longest sequence). We show the mean and standard deviation
    in a sliding window around each learning curve to illustrate this.}
    \label{fig:synthetic_training}
  \end{figure}

>>>>

We investigate this idea further and deliver the following contributions:
\begin{itemize}
\item{We classify RNNs which satisfy the conditions above, and show that many
    RNNs used in practice such as the Quasi-RNNs (QRNNs) introduced by
    \citet{bradbury2017quasi} are contained in this class.}
\item{We provide an implementation of the parallel linear recurrence algorithm
    as a CUDA kernel, and show that it speeds up training of QRNN and
    \citet{lei2017}'s Simple Recurrent Unit (SRU) architectures by factors of up
    to 9x.}
\item{We describe how several recent linear RNNs can be described as
    linear surrogates for non-linear architectures. We introduce a linear
    surrogate for the LSTM and show that we are able to train it
    with a speedup of 5-10x compared to the CuDNN LSTM when we use the
    parallel linear recurrence algorithm.}
\end{itemize}

@@

\subsection{Synthetic Experiment}
One of the key strengths of the LSTM is that it is capable of dealing with
long-term dependencies. In order to demonstrate that the GILR-LSTM is also able
to handle long-term dependencies we tackle a canonical example of inference over
many time steps from \citet{hochreiter1997long}.  We show that in fact the
GILR-LSTM is able to outperform the CuDNN LSTM and extend to sequence lengths
orders of magnitude longer than dealt with previously. The input consists of
sequences of length $n$ where for $n > 0$ each element is a randomly chosen
one-hot vector \(x\) in $p$-dimensional space.  The first vector in each
sequence, \(x_0\), is always either \((1, 0, \ldots, 0)\) or \((-1, 0, \ldots,
0)\). The sequential model must read in an entire sequence and then output the
sign of the first sequence element.  This sequence classification problem
requires remembering the first element over the length of the sequence, and
early RNNs struggled with this for \(p\) as small as a few dozen. In the
original formulation of the problem (dealing in the regime with around one
hundred timesteps), the dimensionality of the input \(p\) is set equal to
\(n\). Since this would make the size of the input data grow impractically large
as \(\mathcal{O}(n^2)\) for long sequences, we fix \(p = 128\) as we vary \(n\).
\begin{figure}[] \centering
\includegraphics[width=12cm]{synthetic_diagram.pdf}
\caption{The structure of the synthetic example and the GILR-LSTM architecture
we used to tackle it. We feed in one-hot unit vectors \(x\) which are chosen
uniformly at random (with replacement). The class is determined by the very
first vector \(x_0\), which has a fixed direction. The sign of \(x_0\)
determines the class. In the diagram, each rounded block indicates a cell of the
RNN, whilst the square indicates a linear unit.}
\label{fig:synthetic_diagram}
\end{figure} We generated sequences for $n$ equal to 1,024, 8,192, and
1,048,576. For each of these we compared a two layer GILR-LSTM with 512 hidden
units to a two layer LSTM with 512 hidden units\footnote{For the longest
sequence length, the number of hidden units was decreased to 64 for both
architectures so that the net could fit in memory.} per layer implemented by
CuDNN.

@@

\begin{table}[]
\label{table:synth-table}
\centering
\caption{Performance of the GILR-LSTM compared to the CuDNN LSTM
  on problem 2b from \citet{hochreiter1997long}. }
\begin{tabular}{@{}lllllll@{}} \toprule Sequence Length &
\multicolumn{2}{c}{\textbf{1,024}} & \multicolumn{2}{c}{\textbf{8,192}} &
\multicolumn{2}{c}{\textbf{1,048,576}} \\ \midrule &CuDNN& GILR &
CuDNN &GILR&CuDNN&GILR\\ \cmidrule(l){2-7} Iterations (1000s) &
                     1.0 %bs=8
                     \(\pm\) 0.4 & 0.55 %bs=8
                                    \(\pm\) 0.04 & 0.44 %bs=8
                                                   \(\pm\) 0.05 & 0.56 %bs=8
                                                                 \(\pm\) 0.16
                                                               &-& 14 \(\pm\) 3 \\
\begin{tabular}[c]{@{}l@{}}Wall time (hours)\end{tabular} &
0.28 \(\pm\) 0.08 & 0.031 \(\pm\) 0.002 & 0.58 \(\pm\) 0.06 & 0.10 \(\pm\) 0.03
&-& 9.7 \(\pm\) 1.7 \\ \bottomrule
\end{tabular}
\end{table}

The results as show in table \ref{table:synth-table} illustrate that the GILR-LSTM is able to converge between 6 and 10
times faster than the CuDNN LSTM. This is somewhat surprising
given the LSTM was specifically constructed for problems of this sort, and the
CuDNN LSTM implementation is highly optimized (to the extent that
the monolithic interface it exposes is difficult to modify or extend). The
GILR-LSTM is implemented entirely in standard TensorFlow with the exception of
using the new linear recurrence op instead of a TensorFlow symbolic loop.
Convergence of the GILR-LSTM models
leads to the conclusion that the non-linearities present in LSTM are
not necessary for solving this instance of the long-term dependency problem.
The time to convergence further leads to the conclusion that inclusion of a
non-linearity at every step incurs a significant training time slowdown.
Furthermore, the GILR-LSTM is able to learn to carry dependencies over a one
million element sequence. As far as we know, this one million step sequence
experiment is the longest sequential learning problem to be handled by neural
networks to date.

\begin{figure}
  \centering
  \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./1k_synthetic_new.png}
  \end{minipage}%
    \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./8k_synthetic_new.png}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \includegraphics[width=1.0\textwidth]{./1m_synthetic_new.png}
  \end{minipage}
  \caption{Learning curves for GILR-LSTM and CuDNN LSTM architectures for various
    sequence lengths. Each plot includes learning curves from multiple independent
    runs. The shaded area around each learning curve shows the rolling variance
    of the loss over minibatches within the run.
}
    \label{fig:synthetic_training}
  \end{figure}
